---
title: "Precision recall for AMP proportions in a genome"
output:
   github_document:
     pandoc_args: "--webtex"
     toc: true
     number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


```{r, echo=FALSE}
library(tidyverse)
library(caret)
library(patchwork)
library(viridis)

source("scripts/calc_cm_metrics.R")
set.seed(6)
```

To start, we read in ampir's v_0.1 trained model as well as the test dataset
```{r}
ampir_v0.1 <- readRDS("data/ampir_0.1.0_data/svm_Radial98_final.rds")
features98Test <- readRDS("data/ampir_0.1.0_data//features98TestNov19.rds")
```

We then use the ampir model to predict the proteins in the test set. We then add the actual cases to the prediction outcome.
```{r}
test_pred_prob <- predict(ampir_v0.1, features98Test, type = "prob")

ampir_prob_data <- test_pred_prob %>%
  add_column(Label = features98Test$Label) %>%
  rename(Neg = Bg) %>%
  rename(prob_AMP = Tg) %>%
  mutate(Label = ifelse(Label == "Tg", "Pos", "Neg"))

```

# Initial look at the precision and recall curves for a low AMP proportion

To create an idea what to set the probability threshold to for AMP prediction in genomes, the precision and recall (AKA sensitivity) curves plot over a range of probability values was constructed using a dataset where AMPs only comprise approximately 1% (10/996).`

First a random 10 true AMPs from the predicted proteins in the ampir v0.1 test set, plus all the true background proteins, were isolated.

```{r}
ampir_prob_data_bg <- ampir_prob_data[grep("Neg", ampir_prob_data$Label),]

ampir_prob_data_tg_10 <- ampir_prob_data[sample(grep("Pos", ampir_prob_data$Label), 10),]

ampir_prob_data_bg_tg10 <- rbind(ampir_prob_data_tg_10, ampir_prob_data_bg)
```

The function `calc_cm_metrics` was then used to calculate a confusion matrix and some associated performance metrics over a range of predicted probability values. The original output of the function is a vector. To use multiple probability threshold, `map_df` is used for iteration and to convert the output to a dataframe. It is then converted to a long format focusing on the precision and recall columns using `pivot_longer` (previously `gather`)

```{r}
pr_curve_sample <- map_df(seq(0.01, 0.99, 0.01), calc_cm_metrics, ampir_prob_data_bg_tg10)

pr_curve_sample_long <- pivot_longer(pr_curve_sample, cols = c("Precision", "Recall"), names_to = "metric", values_to = "value")
```

### Plot 

```{r, echo=FALSE}
ggplot(pr_curve_sample_long, aes(x=p_threshold, y=value)) +
  geom_line(aes(color = metric)) +
  scale_x_continuous(breaks=c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  scale_colour_manual(values = c("blueviolet", "forestgreen"),
                      labels = c("Precision", "Recall")) +
  labs(x = "Probability threshold", y = "", color = "") +
  theme(legend.position = c(0.23, 0.53),
        legend.key = element_rect(fill = "white"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")) +
  guides(color = guide_legend(reverse=TRUE))
```

**Figure 1.1:** The precision and recall curves over a range of probability thresholds for an AMP proportion of 0.01 within the ampir v.0.1 test set


```{r}
ggsave("figures/tenAMPsinlargebg.png", width = 5, height = 3)
```

0.01 is a very small proportion and the curves are creating "big steps". To smoothen out the curves, 100 random selection of 1% (10 AMPs) were used to average the curves.

# Average precision and recall curves for a low AMP proportion

First `replicate` and `slice_sample` were used to select 10 random AMPs 100 times which were bound together in a table with `rbind`. 
```{r}
ampir_prob_data_tg <- filter(ampir_prob_data, Label == "Pos")


samples_list <- replicate(n = 100, ampir_prob_data_tg %>% 
                             slice_sample(n=10) %>% 
                             rbind(ampir_prob_data_bg), simplify = F)
```

Then a nested loop was required to use `calc_cm_metrics` with a range of probability thresholds over the 100 dataframe samples
```{r}
samples_list_metrics <- lapply(seq_along(samples_list), function(i) {
                               map_df(seq(0.01, 0.99, 0.01),
                                        calc_cm_metrics, 
                                        samples_list[[i]])})
```

Now I have a list with 100 dataframes in it that each contain metrics calculated with `calc_cm_metrics` from 100 dataframes that each contain 10 random AMPs and 996 background proteins.

I want to average the precision and recall columns over the 0.01 - 0.99 p_threshold range.
To do this I bind all the dataframes in the list together into one big dataframe using`bind_rows`. Then I use the `group_by` function to group the samples by the p_threshold and then use `summarise` to create two new columns that contain the calculated average of precision and recall and finally convert it into long format again.

```{r}        
samples_metrics <- bind_rows(samples_list_metrics, .id = "sample")

samples_metrics_averages <- samples_metrics %>% 
                group_by(p_threshold) %>% 
                summarise(Precision_average = mean(Precision), 
                          Recall_average = mean(Recall)) %>%
                pivot_longer(cols = c("Precision_average", "Recall_average"),
                               names_to = "metric",
                               values_to = "value") %>%
                mutate(value = coalesce(value, 1)) #replace 0.99 prob NaN value with 1
```

### Plot

```{r, echo = FALSE}
ggplot(samples_metrics_averages, aes(x=p_threshold, y=value)) +
  geom_line(aes(color = metric)) + 
  scale_x_continuous(breaks=c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  scale_colour_manual(values = c("blueviolet", "forestgreen"),
                      labels = c("Precision", "Recall")) +
  labs(x = "Probability threshold", y = "", color = "") +
  theme(legend.position = c(0.23, 0.53),
        legend.key = element_rect(fill = "white"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")) +
  guides(color = guide_legend(reverse=TRUE))

```

**Figure 1.2:** The precision and recall curves over a range of probability thresholds for an average AMP proportion of 0.01 within the ampir v.0.1 test set

# Theoretical AMP proportion in a genome - alpha 

A new variable, $\alpha$, was introduced to represent the percentage of AMPs in the test set to more easily create precision-recall curves for various AMP proportions. 

## alpha equation and implementation

The calculation for the recall metric for $\alpha$ remains the same but the precision metric has been slightly modified (see below):


$$Precision_{\alpha} = \frac{TP\alpha}{TP\alpha + FP(1-\alpha)}$$

$$Recall_{\alpha} = \frac{TP}{TP + FN}$$

Function that uses the recall-precision metric calculations for any $\alpha$ value
```{r}
calc_precision_recall <- function(df,alpha) {
  df %>% 
  mutate(Recall = Recall) %>% 
  mutate(Precision = TP*alpha / (TP*alpha+FP*(1-alpha))) %>% 
  select(Recall,Precision,p_threshold)
}
```

Calculate metrics from real predicted data
```{r}
ampir_roc_data <- map_df(seq(0.01, 0.99, 0.01), calc_cm_metrics, ampir_prob_data)
```

Use the function for a range of $\alpha$ values and collapse to data frame

```{r}
pr_data <- do.call(rbind,lapply(c(0.01,0.05,0.1,0.5),function(alpha) {
  calc_precision_recall(ampir_roc_data,alpha) %>% add_column(alpha=alpha)
}))
```

### Plot

Plot an explicit axis for `p_threshold` using four different $\alpha$ values. This is useful for choosing the threshold value. It shows that as $\alpha$ gets smaller and smaller the Precision curve shifts so that high values of precision are only achieved for very high `p_threshold` values.

```{r, echo = FALSE}
pr_data_long <- pr_data %>% gather("metric","value",-p_threshold,-alpha)


variable_names <- c("0.01" = "Proportion of AMPs in genome: 0.01",
                    "0.05" = "Proportion of AMPs in genome: 0.05",
                    "0.1" = "Proportion of AMPs in genome: 0.10",
                    "0.5" = "Proportion of AMPs in genome: 0.50")

ggplot(pr_data_long,aes(x=p_threshold,y=value)) + 
  geom_line(aes(colour=metric)) + facet_wrap(~alpha, labeller= as_labeller(variable_names)) +
  labs(x = "Probability threshold", y = "", colour = "") +
  scale_x_continuous(breaks=c(0, 0.50, 1.00)) +
  scale_y_continuous(breaks=c(0, 0.50, 1.00)) +
  scale_colour_manual(values = c("blueviolet", "forestgreen")) +
  theme(legend.position = "left",
        legend.text = element_text(angle = 90, hjust = 0.5),
        legend.margin = margin(0,0,0,8),
        legend.box.margin=margin(-10,-10,-10,-10),
        legend.key = element_rect(fill = "white"),
        legend.key.height = unit(2, "cm"),
        legend.box.spacing = unit(0.1, "cm"),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")) +
  guides(linetype = guide_legend(reverse=TRUE))
```

**Figure 1.2:** The precision and recall curves over a range of probability thresholds for four  different proportion of AMPs in a genome.

```{r, echo = FALSE}
ggsave("figures/alphas.png", width = 7, height=5)
```

## Traditional precision-recall curve

Using a traditional precision-recall curve is useful in the sense that it very clearly shows the tradeoff between the two. A useful way to think of Precision is that it defines the "Purity" of our predicted set of AMPs whereas the Sensitivity or Recall defines the "Completeness" of the predicted AMP set. We want to choose the p_threshold so that there is a balance or Purity and Completeness. When $\alpha$ is high this is easy to do, but when it is low it becomes a very difficult tradeoff.

### Plot

```{r, echo=FALSE}
ggplot(pr_data, aes(x=Recall, y=Precision)) +
  geom_line(aes(group=as.factor(alpha), colour = as.factor(alpha))) + 
  scale_color_viridis(discrete = TRUE) +
  theme(legend.key = element_rect(fill = "white"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")) +
  labs(colour = "alpha") +
   guides(color = guide_legend(reverse=TRUE))
```

**Figure 1.3:** A traditional precision-recall curve depicting various alpha values that represent different proportions of AMPs in a genome


```{r, echo=FALSE}
ggsave("figures/alpha_gradient.png", height = 5, width = 7)
```

## Comparing the theoretical 0.01 alpha value with the averaged 0.01 AMP proportion from a real test set

Extract 1% alpha value
```{r}
pr_data_alpha1_long <- pr_data_long %>% filter(alpha == 0.01)
```

### Plot

The precision recall curves plot for $\alpha$ = 0.01 is extremely similar to the average curve of 0.01 AMPs in the ampir v0.1 test set. This was as expected and shows the $\alpha$ variable is a valid depiction as the proportion of AMPs in a test set.

```{r, echo = FALSE}
average_plot1perc <- ggplot(filter(samples_metrics_averages, p_threshold >=0.5), aes(x=p_threshold, y=value)) +
  geom_line(aes(color = metric)) + 
  scale_x_continuous(breaks=c(0, 0.50, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  scale_colour_manual(values = c("blueviolet", "forestgreen"),
                      labels = c("Precision", "Recall")) +
  labs(x = "", y = "", color = "") +
  theme(legend.position = c(0.23, 0.53),
        legend.key = element_rect(fill = "white"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")) +
  guides(color = guide_legend(reverse=TRUE)) +
  ggtitle("A")

alpha_001plot <- ggplot(filter(pr_data_alpha1_long, p_threshold >= 0.5), aes(x=p_threshold, y=value)) +  geom_line(aes(colour = metric)) + 
  labs(x = "Probability threshold", y = "", colour = "") +
  scale_colour_manual(values = c("blueviolet", "forestgreen")) +
  scale_x_continuous(breaks=c(0,0.50, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  theme(legend.position = c(0.2, 0.5),
        panel.background = element_blank(),
        legend.key=element_blank(),
        axis.line = element_line(colour = "grey")) +
  guides(colour = guide_legend(reverse = TRUE)) +
  ggtitle("B")

average1percvsalpha1pc <- average_plot1perc / alpha_001plot

average1percvsalpha1pc
```

**Figure 1.4:** The precision and recall curves over a range of probability thresholds for A: an average AMP proportion of 0.01 within the ampir v.0.1 test set and B: for alpha, a theoretical proportion of AMPs within a genome, equal to 0.01.


```{r, echo = FALSE}
ggsave("figures/alpha_vs_average.png", average1percvsalpha1pc)
```

