---
title: "Benchmarking in an 'omics-scanning context"
output: 
  github_document:
    pandoc_args: "--webtex"

bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r, echo = FALSE}
library(tidyverse)
library(ampir)
library(patchwork)
```

# Preparing Prediction Data

We benchmarked each of the AMP predictors against the following datasets:

- A balanced test dataset as close as possible to the original test data used by the authors of the predictor
- An unbalanced ($\alpha=0.1$) dataset created as a random subsample of the original test data used by the authors
- The reference proteome of *Homo sapiens* (human), obtained from Uniprot, proteome ID: up000005640
- The reference proteome of *Arabidopsis thaliana* (mouse-ear cress), obtained from Uniprot, proteome ID: up000006548

Both Uniprot datasets were downloaded from UniProt proteomes (accessed 23 Jan 2021). In order to ensure that the same data could be used across all predictors we first preprocessed these proteomes to comply with input requirements of all predictors. Specifically we:

- Removed all non-standard amino acids
- Removed all peptides shorter than 10 amino acids

```{r}
reference_proteomes <- c("data/proteomes/uniprot-proteome UP000005640.tab","data/proteomes/uniprot-proteome UP000006548.tab") %>% 
  map_dfr( ~ read_tsv(.x, show_col_types = FALSE) %>% mutate(Label = case_when(str_detect(`Keyword ID`, "KW-0929") ~ "Pos", TRUE ~ "Neg"))) %>% 
  filter(grepl('^[ARNDCEQGHILKMFPSTWYV]+$', Sequence)) %>% 
  filter(nchar(Sequence) >= 10)
```


```{r, eval=FALSE}
# Write fasta data for each organism for later use when benchmarking predictors
#
reference_proteomes %>% 
  filter(Organism=="Arabidopsis thaliana (Mouse-ear cress)") %>% 
  select(`Entry name`, Sequence) %>% 
  df_to_faa("data/proteomes/arabidopsis_benchmark_data.fasta")

reference_proteomes %>% 
  filter(Organism=="Homo sapiens (Human)") %>% 
  select(`Entry name`, Sequence) %>% 
  df_to_faa("data/proteomes/human_benchmark_data.fasta")
```

### ampir

As shown in Figure 1.1 the training data for `ampir` includes precursor protein sequences for both Human and Arabidopsis.  In order to avoid calculating inflated performance metrics for `ampir` we therefore retrained its model after first removing any *H. sapiens* or *A. thaliana* proteins. We report three separate benchmarks for `ampir` (v1.1) as follows:

- `ampir_precursor` : ampir using the unmodified precursor model
- `ampir_prec_nb` : ampir using the precursor model with Human and Arabidopsis proteins removed
- `ampir_mature` : ampir using mature peptides


```{r}
if ( file.exists("data/prediction_results/ampir/ampirv1.1_refproteomes_predictions.rds")){
  ampir_v1.1_proteome_predictions <- readRDS("data/prediction_results/ampir/ampirv1.1_refproteomes_predictions.rds")
} else {

  ampirv1.1_precursor_imbal <- readRDS("data/ampir_v1.1.0_data/tuned_precursor_imbal_full.rds")
  ampirv1.1_mature <- readRDS("data/ampir_v1.1.0_data/tuned_mature_full.rds")
  ampir_v1.1_precursor_imbal_nobench <- readRDS("data/ampir_v1.1.0_data/tuned_precursor_imbal_nobench.rds")

  ampir_models <- list("ampir_v1.1_precursor_imbal"=ampirv1.1_precursor_imbal,
                       "ampir_v1.1_mature"=ampirv1.1_mature,
                       "ampir_v1.1_precursor_imbal_nobench"=ampir_v1.1_precursor_imbal_nobench)

  run_ampir_models_for_proteome <- function(organism_name){
    names(ampir_models) %>% map_dfr( ~ reference_proteomes %>% 
                                          filter(Organism==organism_name) %>% 
                                          select(`Entry name`, Sequence) %>% 
                                          as.data.frame() %>% 
                                          predict_amps(n_cores = 8, model = ampir_models[[.x]]) %>% 
                                          add_column(Model = .x)
                                          )
  }
  
  # These are quite time consuming (~5 minutes on a Mac Mini with 8 cores)
  cress_pred <- run_ampir_models_for_proteome("Arabidopsis thaliana (Mouse-ear cress)")
  human_pred <- run_ampir_models_for_proteome("Homo sapiens (Human)")


  ampir_v1.1_proteome_predictions <- rbind(cress_pred,human_pred) %>% 
    left_join(reference_proteomes, by = "Entry name") %>% 
    select(ID = `Entry name`, prob_AMP, Organism, Label, Model)
  
  saveRDS(ampir_v1.1_proteome_predictions, "data/prediction_results/ampir/ampirv1.1_refproteomes_predictions.rds")
}
```

The training and test data for the AMP predictors are outlined in [01_data_comparison.md](01_data_comparison.md), Table 1.1. Predictors were run in January 2021.

## Antimicrobial Peptide Scanner vr. 2 ()

AMPscanner vr.2 Feb2020 model was used via the [AMPscanner webserver](https://www.dveltri.com/ascan/v2/ascan.html)

```{r}
ampscanner_file_paths <- c(list.files("data/prediction_results/ampscanner_v2", pattern="*.csv",full.names = T))

ampscan_genome_bench <- list.files("data/prediction_results/ampscanner_v2", pattern="[ha][ro][am].*.csv",full.names = T) %>% 
  map_dfr( ~ read_csv(.x, show_col_types = FALSE) %>% separate(SeqID,into = c("database","Entry","Entry name"),sep = "\\|") ) %>% 
  left_join(reference_proteomes,by="Entry name") %>% 
  select(ID = `Entry name`, prob_AMP = Prediction_Probability, dataset=Organism, Label)

ampscan_pub_bench <- list.files("data/prediction_results/ampscanner_v2", pattern="amps_.*.csv",full.names = T) %>% 
  map_dfr( ~ read_csv(.x, show_col_types = FALSE) %>% add_column(Label = ifelse(grepl("nonamps",.x),"Neg","Pos"))) %>% 
  add_column(dataset = "published") %>% 
  select(ID = SeqID, prob_AMP = Prediction_Probability, dataset, Label)

ampscan_bench <- rbind(ampscan_genome_bench,ampscan_pub_bench) %>% add_column(Model = "AMPscanner_v2")
```

## amPEP

amPEP MATLAB source code was accessed from [sourceforge](https://sourceforge.net/projects/axpep/files/AmPEP_MATLAB_code/) in January 2021.

```{r}
ampep_genome_bench <- c("data/prediction_results/ampep/arab_proteome_standardaa_ampep.txt","data/prediction_results/ampep/homo_proteome_standardaa_ampep.txt") %>% 
  map_dfr( ~ read_csv(.x, show_col_types = FALSE) %>% 
             separate(Row, into = c("database","Entry","Entry name"),sep = "\\|") %>% 
             left_join(reference_proteomes, by = "Entry name") %>% 
             select(ID = `Entry name`, prob_AMP = score, dataset=Organism, Label)
           )

ampep_pub_bench <- read_csv("data/prediction_results/ampep/xiao_independent_ampep.txt", show_col_types = FALSE) %>% 
  mutate(Label = ifelse(grepl(Row,pattern = "^AP"), "Pos", "Neg")) %>% 
  add_column(dataset="published") %>% 
  select(ID = Row, prob_AMP = score, dataset, Label)

ampep_bench <- rbind(ampep_genome_bench,ampep_pub_bench) %>%
  add_column(Model = "amPEP")
```


## amPEPpy

amPEPpy v.1.0 was run according to the example provided on the [amPEPpy GitHub repository](https://github.com/tlawrence3/amPEPpy) in a Conda (v. 4.10.0) environment (commands shown below). 


```bash
ampep predict -m pretrained_models/amPEP.model -i ../cache/arab_proteome_standardaa.fasta -o arab_results.tsv --seed 2012
ampep predict -m pretrained_models/amPEP.model -i ../cache/homo_proteome_standardaa.fasta -o homo_results.tsv --seed 2012
```

```{r}
ampeppy_genome_bench <- list.files("data/prediction_results/amPEPpy", pattern="[ah][or][am].*.tsv",full.names = T) %>% 
  map_dfr( ~ read_tsv(.x, show_col_types = FALSE) %>% 
             separate(seq_id, into = c(NA, NA,"Entry name"),sep = "\\|") %>%
             left_join(reference_proteomes, by = "Entry name") %>% 
             select(ID = `Entry name`, prob_AMP = probability_AMP, dataset = Organism, Label))
  
ampeppy_pub_bench <-  read_tsv("data/prediction_results/amPEPpy/xiao_independent_ampeppy_results.tsv", show_col_types = FALSE) %>% 
  mutate(Label = ifelse(grepl(seq_id,pattern = "^AP"), "Pos", "Neg")) %>% 
  add_column(dataset="published") %>% 
  select(ID = seq_id, prob_AMP = probability_AMP, dataset, Label)

ampeppy_bench <- rbind(ampeppy_genome_bench,ampeppy_pub_bench) %>% add_column(Model = "amPEPpy")
```

## AMPlify

AMPlify v1.0.0 was installed as the latest release from [AMPlify's repository](https://github.com/bcgsc/AMPlify). However, when using AMPlify to predict proteins an error arose, related to the software itself, and therefore unfortunately could not be included in the benchmark. An issue was raised with details about this error on the [AMPlify's issue page](https://github.com/bcgsc/AMPlify/issues/1) on 10/02/2021. 

Update: when looking at the commit history, a [commit](https://github.com/bcgsc/AMPlify/commit/2c44491b7a8f4815b154307f5b250f991942935a) was found which looked like it would fix the issue. After installing the released version, I changed the source code for the software according to the previously mentioned commit, and the software worked after that. 

AMPlify is optimised for protein sequences that are 200 AA or less and do not allow input data to contain sequences larger than 200 amino acids. Therefore, only proteins that were =< 200 AA were kept in the proteomes (9,027 for *A. thaliana* from 39,340 standard AA sequences and 33,584 for *H. sapiens* from 67,484 standard AA sequences)


```{bash, eval = FALSE}
cat data/prediction_results/AMPlify1.0.0/AMPlify_results_20210405153447_Athaliana.txt | sed 's/\r//' | awk '$0~/^Sequence ID:/{spid=$3};$0~/^Score:/{printf("%s\t%s\n",spid,$2)}' > data/prediction_results/AMPlify1.0.0/AMPlify_Athaliana.txt 

cat data/prediction_results/AMPlify1.0.0/AMPlify_results_20210405182202_Hsapiens.txt | sed 's/\r//' | awk '$0~/^Sequence ID:/{spid=$3};$0~/^Score:/{printf("%s\t%s\n",spid,$2)}' > data/prediction_results/AMPlify1.0.0/AMPlify_Hsapiens.txt 

cat data/prediction_results/AMPlify1.0.0/AMPlify_results_20210531101008amps.txt | sed 's/\r//' | awk '$0~/^Sequence ID:/{spid=$3};$0~/^Score:/{printf("%s\t%s\n",spid,$2)}' > data/prediction_results/AMPlify1.0.0/AMPlify_amps.txt 

cat data/prediction_results/AMPlify1.0.0/AMPlify_results_20210531101827nonamps.txt | sed 's/\r//' | awk '$0~/^Sequence ID:/{spid=$3};$0~/^Score:/{printf("%s\t%s\n",spid,$2)}' > data/prediction_results/AMPlify1.0.0/AMPlify_nonamps.txt 
```

```{r}
amplify_genome_bench <- c("data/prediction_results/AMPlify1.0.0/AMPlify_Athaliana.txt","data/prediction_results/AMPlify1.0.0/AMPlify_Hsapiens.txt") %>% 
  map_dfr( ~ read_delim(.x,delim = "\t", col_names = c("Entry name", "prob_AMP"), show_col_types = FALSE) %>% 
             separate(`Entry name`, into = c(NA, NA,"Entry name"),sep = "\\|") %>%
             left_join(reference_proteomes, by = "Entry name") %>% 
             select(ID = `Entry name`, prob_AMP, dataset=Organism, Label)
             )

amplify_pub_bench <- c("data/prediction_results/AMPlify1.0.0/AMPlify_amps.txt","data/prediction_results/AMPlify1.0.0/AMPlify_nonamps.txt") %>% 
  map_dfr( ~ read_delim(.x,delim = "\t", col_names = c("Entry name", "prob_AMP"), show_col_types = FALSE) %>% 
             add_column(Label = ifelse(grepl("nonamps",.x),"Neg","Pos")) %>% 
             add_column(dataset="published") %>% 
             select(ID = `Entry name`, prob_AMP, dataset, Label)
  )

amplify_bench <- rbind(amplify_genome_bench,amplify_pub_bench) %>% 
  add_column(Model = "AMPlify")

```


## AMPgram


In order to significantly speed up the progress of using AmpGram v. 1.0 on the *A. thaliana* and *H. sapiens* proteomes, high performance computing (HPC) scheduler (PBS) with job arrays submissions were used (with R v. 4.0.3). To accomplish this, first, both proteomes were split into FASTA files containing 100 protein sequences (394 FASTA files for *A. thaliana* and 675 for *H. sapiens*) using the scripts `scripts/subset_arab_file.zsh` and `subset_homo_proteome.zsh`. Per job, approximately 100 subjobs which referenced to 100 FASTA files were used. See `scripts/runampgram_h1.sh` and `scripts/runampgram_h1.R` as example scripts used for the first 100 subjobs/FASTA files from the *H. sapiens* proteome. 

Initially when the jobs were submitted, various subjobs failed (three for *A. thaliana* and 62 for *H. sapiens*). After examining the [source code for AmpGram](https://github.com/michbur/AmpGram/blob/master/R/utils.R), it appears that AmpGram does not support sequences less than 10 amino acids long. In the *A. thaliana* proteome (with non standard amino acids removed) there were three proteins less than 10 amino acids long and in *H. sapiens* there were 268. Therefore these sequences were not included in the prediction analysis from AmpGram.

*AmpGram's prediction results is a list for each protein prediction that is subdivided into three sublists. For this benchmark analysis, only the third sublist is relevant as this contains the probability score for that protein. A function was written to easily extract this probability score and associated protein name from each output file. 

```{r}
gimme_ampgram_predictions <- function(filepath) {
  
  ampgram_list <- readRDS(filepath)

  amgram_probandnamelist <- lapply(ampgram_list, '[[', 3)

  bind_rows(amgram_probandnamelist, .id = "ID")
}
```


```{r}
if (file.exists("cache/ampgram_genome_bench.rds")) {
  ampgram_genome_bench <- readRDS("cache/ampgram_genome_bench.rds")
} else {
  
ampgram_filepaths <- list.files(c("data/prediction_results/ampgram/homo", "data/prediction_results/ampgram/homo/leftovers", "data/prediction_results/ampgram/arab"), pattern="*.rds",full.names = T)

ampgram_predictions <- map_df(ampgram_filepaths, gimme_ampgram_predictions)
  
ampgram_genome_bench <- ampgram_predictions %>%
  separate(ID, into = c(NA, NA,"Entrynamewdescr"),sep = "\\|") %>%
  separate(Entrynamewdescr, into = c("Entry name", NA), sep = "\t", fill = "left") %>%
  left_join(reference_proteomes, by = "Entry name") %>% 
  select(ID = `Entry name`, prob_AMP = "TRUE", Organism, Label) %>% 
  add_column(Model = "AmpGram")

  saveRDS(ampgram_genome_bench, "cache/ampgram_genome_bench.rds")
}
```



### Calculating performance metrics - ROC curves

A function, `get_genome_roc.R` was written to use `calc_cm_metrics.R` to calculate performance metrics over a range of predicted probability (0 - 0.99) values, which include metrics necessary to construct ROC curves (false positive rate and true positive rate) 

```{r}
source("scripts/calc_cm_metrics.R")

organisms = c("Homo sapiens (Human)","Arabidopsis thaliana (Mouse-ear cress)")

get_genome_roc <- function(data, model_name){
  do.call(rbind,lapply(organisms,function(org){ 
    map_df(c(seq(0.01, 0.99, 0.01),seq(0.99, 0.990, 0.001)), calc_cm_metrics, data %>% filter(Organism==org)) %>%
    add_column(Organism = org)
  })) %>%   
  add_column(Model = model_name)
}
```

To use `get_genome_roc.R` on ampir data, an additional loop had to be implemented as in this case, ampir is subdivided into three different models and therefore metric calculations needed to be done three different times, one for each model.

```{r}
ampir_genome_roc <- do.call(rbind,lapply(c("ampir_v1.1_precursor_imbal","ampir_v1.1_mature", "ampir_v1.1_precursor_imbal_nobench"),function(meth){
    get_genome_roc(ampir_v1.1_proteome_predictions %>% filter(Model==meth),meth)}))
```


```{r}
ampscanner_roc <- get_genome_roc(ampscan_genome_bench %>% rename(Organism=dataset), "AMPscanner_v2")

ampep_roc <- get_genome_roc(ampep_genome_bench %>% rename(Organism=dataset), "amPEP")

ampgram_roc <- get_genome_roc(ampgram_genome_bench, "AmpGram")

ampeppy_roc <- get_genome_roc(ampeppy_genome_bench %>% rename(Organism=dataset), "amPEPpy")

amplify_roc <- get_genome_roc(amplify_genome_bench %>% rename(Organism=dataset), "AMPlify")
```

```{r}
proteome_rocs <- rbind(ampir_genome_roc, ampscanner_roc, ampep_roc, ampgram_roc, ampeppy_roc, amplify_roc)
```

## Plots

ROC curves, based on the false and true positive rates, and subsequent AUROCs are often used to evaluate model performance but these can be misleadingly confident when used on a dataset that is highly imbalanced, i.e. where one class heavily outweighs the other [Davis & Goadrich 2006](https://doi.org/10.1145/1143844.1143874). A more accurate alternative would be to use precision and recall (PR) curves as these focus on the proportion of actual true positives within the positive predictions [Saito & Rehmsmeier 2015](https://dx.doi.org/10.1371%2Fjournal.pone.0118432), rather than including the true negatives, as the false positive rates in the ROC curves do. As the proportion of AMPs in a genome is extremely low, the AMP prediction models would have to perform on a highly imbalanced dataset and therefore the precision recall curves are additionally used in this study. ROC curves are evaluated on their shape, generally curves that arc closer to the top left corner (similar to a capital gamma Γ shape), as well as their AUC values. The AUC values range between 0.0 and 1.0 where 0.0 the model cannot tell the classes apart at all and 1.0 the model is able to distinguish between classes perfectly. Generally, an AUC of 0.5 already indicates that the model struggles to differentiate between classes. On the ROC curve plot, an AUC of 0.5 makes a diagonal line from the bottom left to the top right corner. The perfect PR curve is like a mirror image of the perfect ROC curve; it bends at the top right corner, which refers to the model performing with 100% recall and precision. Therefore, the more the PR curves bend toward the top right corner, the better the model is. When comparing multiple curves on the same plot, the curve that is above another curve, is generally assumed to perform better.

Figure 3.1 shows both the ROC curves (top row) and the PR curves (bottom row) for the prediction results from various AMP predictors on the *H. sapiens* and *A. thaliana* proteomes. It is clear that the ROC curves overall show a better performance compared to the PR curves which corroborates the findings of David and Goadrich (2006) and Saito and Rehsmeier (2015). The AUC values for the ROC curves range between 0.16 - 0.99 for *A. thaliana* and 0.43 - 0.94 for *H. sapiens*. The AUC values for the PR curves are between 0.004 - 0.83 for *A. thaliana* and 0.001 - 0.30 for *H. sapiens*, which are overall much lower compared to the AUROC values. Interestingly, all models, with exception to amPEP, had higher AUC values for both ROC and PR curves for *A thaliana*. This may indicate the models were better at detecting AMPs in *A. thaliana* compared to *H. sapiens*. The ampir precursor model had the highest PRAUC value on the *A. thaliana* proteome (PRAUC: 0.83). However, the remaining PRAUC values (for both proteomes) are below 0.34. Therefore, according to the AUPRC metric, which has been stated to be more informative on imbalanced datasets, none of the models (save perhaps the ampir precursor model on *A. thaliana*) were skilled enough to detect AMPs in the *H. sapiens* and *A. thaliana* proteomes.

```{r, echo = FALSE, fig.width=10, fig.height=7}
rocplot <- ggplot(proteome_rocs) + 
  geom_line(aes(x = FPR, y = Recall, colour = Model)) + 
  facet_wrap(~Organism) +
  labs(x= "False positive rate", y = "True positive rate", colour = "") +
  theme_classic() +
  theme(legend.position = "bottom",
        strip.background = element_blank(),
        strip.text = element_text(face = "italic",
                                  size = 10)) +
 scale_colour_manual(breaks = c("ampir_v1.1_precursor_imbal","ampir_v1.1_mature", "ampir_v1.1_precursor_imbal_nobench",
                                "AMPscanner_v2", "amPEP", "AmpGram",
                                "amPEPpy", "AMPlify"),
                     labels = c("ampir_precursor\nH .94 A .99","ampir_mature\n H .74 A .97","ampir_prec_nb\nH .85 A .95",
                                "AMPscanner\nH .79 A .92", "amPEP\nH .43 A .16", "AmpGram\nH .81 A .86",
                                "amPEPpy\nH .49 A .24","AMPlify\n H .67 A .62"),
                     values = c("blueviolet", "goldenrod2", "black","cyan", "green",
                                "darkorange3", "grey50", "deeppink")) +
 guides(colour = guide_legend(nrow = 1))


prplot <- ggplot(proteome_rocs) + 
  geom_line(aes(x = Recall, y = Precision, colour = Model)) + 
  facet_wrap(~Organism, scales = "free") +
  labs(x= "Recall", y = "Precision", colour = "") +
  theme_classic() +
  theme(legend.position = "bottom",
        strip.background = element_blank(),
        strip.text = element_blank()) +
 scale_colour_manual(breaks = c("ampir_v1.1_precursor_imbal","ampir_v1.1_mature",
                                "ampir_v1.1_precursor_imbal_nobench",
                                "AMPscanner_v2", "amPEP", "AmpGram", "amPEPpy", "AMPlify"),
                     labels = c("ampir_precursor\nH .30 A .83","ampir_mature\nH .004 A. 15","ampir_prec_nb\nH .11 A. 34",
                                "AMPscanner\nH .006 A .09", "amPEP\nH .001 A .004", "AmpGram\nH .01 A .14",
                                "amPEPpy\nH .001 A .005","AMPlify\n H .004 A .05"),
                     values = c("blueviolet", "goldenrod2", "black","cyan", "green",
                                "darkorange3", "grey50", "deeppink")) +
   guides(colour = guide_legend(nrow = 1))

rocplot / prplot  + plot_annotation(tag_levels = 'A')
```

**Figure 3.1:** Performance of various AMP predictors in classifying whole proteome data for *Homo sapiens* and *Arabidopsis thaliana*. Performance is shown as ROC curves (top row) and precision-recall curves (second row). 
H refers to *Homo sapiens* and A refers to *Arabidopsis thaliana*. The numbers that follow are the respective AUC values for either the ROC or PR curve. 

```{r, echo = FALSE, eval = FALSE}
rocandprplot <- rocplot / prplot + plot_annotation(tag_levels = 'A')

ggsave(plot = rocandprplot, filename = "figures/proteomesrocandpr.png", width = 25, height = 18, units = "cm")
```


Similar to how the ROC curves were calculated, a function, `get_metrics.R`, was written to calculate performance metrics, which include the area under the curve (AUC) for both the ROC and Precision-recall curves, of the various AMP models tested on the *H. sapiens* and *A. thaliana* proteomes.
```{r}
source("scripts/calculate_model_metrics.R")

get_metrics <- function(bench_data, model_name) {
  do.call(rbind,lapply(c("Homo sapiens (Human)","Arabidopsis thaliana (Mouse-ear cress)"),function(org){
    calculate_model_metrics(bench_data %>% filter(Organism==org)) %>%
      add_column(Organism = org) %>% 
      add_column(Model = model_name)
    }))
}

ampir_metrics <- do.call(rbind, lapply(c("ampir_v1.1_precursor_imbal","ampir_v1.1_mature", "ampir_v1.1_precursor_imbal_nobench"),function(meth) {
  get_metrics(ampir_v1.1_proteome_predictions %>% filter(Model==meth), model_name = meth)
}))

ampscan_metrics <- get_metrics(ampscan_genome_bench %>% rename(Organism=dataset), "AMPscanner_v2")
ampep_metrics <- get_metrics(ampep_genome_bench %>% rename(Organism=dataset), "amPEP")
ampgram_metrics <- get_metrics(ampgram_genome_bench, "AmpGram")
ampeppy_metrics <- get_metrics(ampeppy_genome_bench %>% rename(Organism=dataset), "amPEPpy")
amplify_metrics <- get_metrics(amplify_genome_bench %>% rename(Organism=dataset), "AMPlify")

proteome_metrics <- rbind(ampir_metrics, ampscan_metrics, ampep_metrics, ampgram_metrics, ampeppy_metrics, amplify_metrics) %>% mutate(Organism = case_when(str_detect(Organism, "Homo") ~ "H. sapiens", TRUE ~ "A. thaliana"))
```

**Table 3.1:** Performance metrics of various predictors on the proteomes of *Homo sapiens* and *Arabidopsis thaliana*

```{r, echo = FALSE}
knitr::kable(proteome_metrics, digits = 2)
```


```{r, fig.height=10, fig.width=9}
proteome_metrics_long <- proteome_metrics %>% 
  select(Recall, Precision, MCC, AUROC, AUPRC, Organism, Model) %>% 
  pivot_longer(cols = c(-Organism, -Model)) %>% 
  mutate(Model = factor(Model, levels = c("ampir_v1.1_precursor_imbal", "ampir_v1.1_mature", "ampir_v1.1_precursor_imbal_nobench", "AMPscanner_v2", "amPEP", "AmpGram", "amPEPpy", "AMPlify")))
  
model_labels <- c("ampir_precursor", "ampir_mature","ampir_prec_nb", "AMPscanner","amPEP", "AmpGram", "amPEPpy", "AMPlify")
model_col_values <- c("blueviolet", "goldenrod2", "black", "cyan", "green","darkorange3","grey50", "deeppink")

ggplot(proteome_metrics_long, aes(x = Organism, y = value)) +
  geom_bar(stat = "identity", aes(fill = Model), position = "dodge") +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  theme_classic() +
  theme(legend.position = "bottom",
        strip.background = element_rect(colour = "white"),
        strip.text = element_text(face = "bold", size = 10),
        axis.text.x = element_text(face = "italic", size = 10)) +
  scale_fill_manual(breaks = c("ampir_v1.1_precursor_imbal","ampir_v1.1_mature",
                                "ampir_v1.1_precursor_imbal_nobench",
                                "AMPscanner_v2", "amPEP", "AmpGram", "amPEPpy", "AMPlify"),
                     labels = model_labels,
                     values = model_col_values) +
  labs(x = "", fill = "", y = "Performance metric value") +
  guides(fill = guide_legend(nrow = 1))
```

**Figure 3.2:** Performance of various AMP predictors in classifying AMPs in whole proteome data for *Homo sapiens* and *Arabidopsis thaliana*.

```{r}
ggsave("figures/proteome_metrics_groupedbar.png", width = 22.5, height = 20, units = "cm")
```

The metrics overall are really low for the ability of models to predict AMPs in proteomes. However, these metrics may not be a very informative evaluation.

It is important to remember the real life situation and applications of predictive models. The actual frequency of AMPs (true positives) in a proteome is approximately 1%. This is extremely low. To adequately express the real-world performance of predictors on proteomes, the numbers of true and false positives were used, with a focus on the low false positive regime, as this is what matters most in whole proteome scans (Figure 3.2)

```{r, echo=FALSE, fig.width=9, fig.height=5}
proteome_rocs <- proteome_rocs %>% mutate(Organism = case_when(str_detect(Organism, "Homo") ~ "Homo sapiens", TRUE ~ "Arabidopsis thaliana"))

ggplot(proteome_rocs) + 
  geom_line(aes(x = FP, y = TP, colour = Model),size = 1.1) + 
  xlim(0,500) +
  facet_wrap(~Organism, scales = "free_y", nrow = 1) +
  labs(x= "False positives", y = "True positives", colour = "") +
  theme_classic() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "italic"),
        strip.background = element_blank()) +
  scale_colour_manual(breaks = c("ampir_v1.1_precursor_imbal", "ampir_v1.1_mature","ampir_v1.1_precursor_imbal_nobench",
                                 "AMPscanner_v2", "amPEP", "AmpGram", "amPEPpy", "AMPlify"),
                     labels = model_labels,
                     values = model_col_values) +
   guides(colour = guide_legend(nrow = 1))
```

**Figure 3.3:** The ability of various models to predict AMPs in the low false positive regime (<500) in the proteomes of *Arabidopsis thaliana* and *Homo sapiens*. It is scaled so that the limits of the y-axis show the full complement of known AMPs in each genome (294 for *A. thaliana*, 112 for *H. sapiens*), and the limits of the x-axis are restricted to emphasise behaviour in the low false positive (FP) regime (FP < 500). 

```{r, eval=FALSE, echo=FALSE}
ggsave("figures/TPFPcountproteomesXLIM.png", width = 22.3, height = 9 ,units = "cm")
```

Note that AMP scanner isn't visible in Figure 3.3 as it only predicts outside the false positive regime (see Figure 3.4 below which is identical to Figure 3.3 except that it shows the full x-axis)

```{r, echo=FALSE, fig.width=9, fig.height=5}
proteome_rocs <- proteome_rocs %>% mutate(Organism = case_when(str_detect(Organism, "Homo") ~ "Homo sapiens", TRUE ~ "Arabidopsis thaliana"))

ggplot(proteome_rocs) + 
  geom_line(aes(x = FP, y = TP, colour = Model),size = 1.1) + 
  facet_wrap(~Organism, scales = "free", nrow = 1) +
  labs(x= "False positives", y = "True positives", colour = "") +
  theme_classic() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "italic"),
        strip.background = element_blank()) +
  scale_colour_manual(breaks = c("ampir_v1.1_precursor_imbal", "ampir_v1.1_mature","ampir_v1.1_precursor_imbal_nobench",
                                 "AMPscanner_v2", "amPEP", "AmpGram", "amPEPpy", "AMPlify"),
                     labels = model_labels,
                     values = model_col_values) +
   guides(colour = guide_legend(nrow = 1))
```

**Figure 3.4:** The ability of various models to predict AMPs in the proteomes of *Arabidopsis thaliana* and *Homo sapiens* using the full x-axis range.

```{r, eval=FALSE, echo=FALSE}
ggsave("figures/TPFPcountproteomes.png", width = 22.3, height = 9 ,units = "cm")
```
