---
title: "Taxonomic composition of Swiss-Prot AMPs"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

UniProt AMPs were obtained from www.uniprot.org with the keyword "antimicrobial" [KW-0929](https://www.uniprot.org/keywords/KW-0929).
On 14 January 2021 this returned 3,339 Reviewed proteins and 41,477 Unreviewed proteins.

All 3,339 reviewed entries were downloaded as a tab separated file which included the sequence and taxonomic lineage (CLASS and ALL) as additional columns. Note that the CLASS column was not used in the code analysis and therefore is not strictly necessary. However, it helped me as a visual tool to group the organisms together.

```{r, echo=FALSE}
library(tidyverse, quietly = TRUE, warn.conflicts = FALSE)
```

```{r}
swissprot_amps <- read_tsv("data/uniprot-keyword Antimicrobial+[KW-0929] -filtered-reviewed yes.tab") %>%
    rename(Class = `Taxonomic lineage (CLASS)`) %>%
    rename(Taxonomic_lineage = `Taxonomic lineage (all)`)
```

Organisms were grouped according to their taxonomic class as much as possible. However, to simplify, some cases were grouped together by colloquial terms for cases where multiple Class arrangements were present or where using a different taxonomic level was easier, e.g. the order Testudines (which includes the turtles) and order Squamata (which includes lizards and snakes) were combined into the "Reptiles" group. Furthermore, Ascidiacea (which includes sea squirts), Amphioxus (lancelets), Cnidaria (includes jellyfishes), Echinodermata (includes sea stars), Mollusca (includes octopods), Merostomata (horseshoe crabs) and Crustacea (includes crabs) were all grouped into "Marine invertebrates" as their individually known AMP count was low.


```{r}
swissprot_amps_grouped <- swissprot_amps %>%
mutate(Taxonomic_lineage = case_when(
   str_detect(Taxonomic_lineage, "Mammalia") ~ "Mammals",
   str_detect(Taxonomic_lineage, "Insecta") ~ "Insects",
   str_detect(Taxonomic_lineage, "Arachnida") ~ "Arachnids",
   str_detect(Taxonomic_lineage, "Bacteria") ~ "Bacteria",
   str_detect(Taxonomic_lineage, "Viruses") ~ "Bacteriophages",
   str_detect(Taxonomic_lineage, "Amphibia") ~ "Amphibians",
   str_detect(Taxonomic_lineage, "Viridiplantae") ~ "Plants",
   str_detect(Taxonomic_lineage, "Fungi") ~ "Fungi",
   str_detect(Taxonomic_lineage, "Annelida") ~ "Worms",
   str_detect(Taxonomic_lineage, "Trematoda") ~ "Worms",
   str_detect(Taxonomic_lineage, "Nematoda") ~ "Worms",
   str_detect(Taxonomic_lineage, "lamprey") ~ "Fishes",
   str_detect(Taxonomic_lineage, "Actinopteri") ~ "Fishes",
   str_detect(Taxonomic_lineage, "Testudines") ~ "Reptiles",
   str_detect(Taxonomic_lineage, "Squamata") ~ "Reptiles", 
   str_detect(Taxonomic_lineage, "Aves") ~ "Birds", 
   str_detect(Taxonomic_lineage, "Ascidiacea") ~ "Marine invertebrates", 
   str_detect(Taxonomic_lineage, "Amphioxus") ~ "Marine invertebrates", 
   str_detect(Taxonomic_lineage, "Cnidaria") ~ "Marine invertebrates", 
   str_detect(Taxonomic_lineage, "Echinodermata") ~ "Marine invertebrates", 
   str_detect(Taxonomic_lineage, "Mollusca") ~ "Marine invertebrates",
   str_detect(Taxonomic_lineage, "Merostomata") ~ "Marine invertebrates",
   str_detect(Taxonomic_lineage, "Crustacea") ~ "Marine invertebrates",
   str_detect(Taxonomic_lineage, "Amoebozoa") ~ "Amoebae",
   str_detect(Taxonomic_lineage, "Archaea") ~ "Archaea",
   str_detect(Taxonomic_lineage, "Chilopoda") ~ "Centipedes",
                                        TRUE ~ Taxonomic_lineage))
```

```{r}
swissprot_amp_counts <- swissprot_amps_grouped %>%
  group_by(Taxonomic_lineage) %>% 
  summarise(amp_count = n()) %>% 
  mutate(percentage = round(amp_count / sum(amp_count) * 100, digits = 1))
```

Amphibians make up the majority of AMPs in the Swiss-Prot database (~24%), followed by mammals (~19%), arthropods (~20%) and plants (16%). Marine invertebrates, which include several large and diverse taxonomic groups, represented here by cnidarians, echinoderms and crustaceans, comprise only 3.5% of the currently known AMPs. It is clear that there is substantial taxonomic bias in the organism groups that represent the AMPs in Swiss-Prot.

```{r, echo = FALSE}
ggplot(swissprot_amp_counts, aes(x = reorder(Taxonomic_lineage, amp_count), y = amp_count)) +
  geom_col(width = 0.7, fill = "darkslategrey") +
  coord_flip() +
  geom_text(aes(label=paste0(percentage, "%")), size = 2.4, hjust = -0.1) +
  labs(y = "Number of AMPs discovered", x = "Organism group") +
  theme_classic()
```

**Figure 4.1:** Antimicrobial peptides (AMPs) present within groups of organisms as obtained from SwissProt 

```{r, echo = FALSE}
ggsave("figures/swissprot_amps.png",  width = 20, height = 20, units = "cm")
```

