---
title: "Taxonomic bias effect on model performance"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo = FALSE}
library(tidyverse)
library(ampir)
library(patchwork)
set.seed(345)
```

Known AMPs are not evenly distributed and are taxonomically biased. This might influence the ability of a predictor to predict AMPs in an organism that is not represented, or does not have close relatives, in the positive AMP dataset.
To investigate this, a predictive model was trained on deuterostomes and tested on protostomes. Deuterostomia and Protostomia are two well established clades that comprise the bilaterian animals, which split over 500 million years ago [Heger et al. 2020](https://doi.org/10.7554/eLife.45530).  

AMPs (the positive datasets) were obtained from Swiss-Prot (accessed 15 January 2021) using the keyword "Antimicrobial [KW-0929]" and saved as a tab separated file which included the Taxonomic Lineage (all) and the sequence column. Deuterostome and protostome AMPs (1,702 and 712, respectively) were extracted from this file.

```{r}
swissprot_amps <- read_tsv("data/uniprot-keyword Antimicrobial+[KW-0929] -filtered-reviewed yes.tab") %>%
    rename(Taxonomic_lineage = `Taxonomic lineage (all)`) %>%
    rename(Entry_name = `Entry name`) %>%
    mutate(Taxonomic_lineage = case_when(
       str_detect(Taxonomic_lineage, "Deuterostomia") ~ "Deuterostome",
       str_detect(Taxonomic_lineage, "Protostomia") ~ "Protostome",
                                        TRUE ~ Taxonomic_lineage))
```

# extract AMPs 0 - 300 maybe for both superphyla and also match negative dataset to this sequence length to avoid length bias in the model 
```{r}
deuterostome_amps <- filter(swissprot_amps, Taxonomic_lineage == "Deuterostome") %>% select(Entry_name, Sequence)

protostome_amps <- filter(swissprot_amps, Taxonomic_lineage == "Protostome") %>% select(Entry_name, Sequence)
```

Plot the sequence length distribution for Deuterostome and Protostome AMPs
```{r, echo = FALSE}
deut_pos_lengths <- ggplot(filter(swissprot_amps, Taxonomic_lineage == "Deuterostome"), aes(Length)) + 
  geom_histogram()

prot_pos_lengths <- ggplot(filter(swissprot_amps, Taxonomic_lineage == "Protostome"), aes(Length)) + 
  geom_histogram()

deut_pos_lengths + prot_pos_lengths
```

The negative datasets were also obtained from Swiss-Prot using the taxonomy terms Deuterostomia [33511] or Protostomia [33317] and excluded the AMPs by using the NOT "Antimicrobial" keyword.

```{r}
deuterostome_neg <- read_faa("data/uniprot-taxonomy Deuterostomia+[33511] +reviewed yes+NOT+keyword--.fasta") %>%
  mutate(Length = nchar(seq_aa))
protostome_neg <- read_faa("data/uniprot-taxonomy Protostomia+[33317] +reviewed yes+NOT+keyword--.fasta") %>%
  mutate(Length = nchar(seq_aa))
```

```{r}
deut_neg_lengths <- ggplot(deuterostome_neg, aes(Length)) + 
  geom_histogram()

prot_neg_lengths <- ggplot(protostome_neg, aes(Length)) + 
  geom_histogram()

deut_neg_lengths + prot_neg_lengths
```


